\chapter{Model Implementation and Evaluation Tools}
A significant portion of the work for this thesis was dedicated to preprocessing the dataset from the source SNN model, developing the RNN model, and creating an evaluation framework. All source code related to these functionalities is included in the thesis attachment, accompanied by documentation providing comprehensive descriptions of each tool.

Throughout the project, development was managed using a GitHub version control repository, where the most up-to-date version of the thesis source code remains available. The attachment provided with this thesis is a snapshot of our working branch \verb|model_restruct|. Due to file type restrictions and large file size, a small example dataset was removed from the attachment. For access to an example dataset, the latest project version, or to review the development history, please refer to the project's GitHub repository:\footnote{\url{https://github.com/dbeinhauer/mcs-source}}.

It should be noted that a small portion of the codebase was contributed by another student, Richard Kraus, who worked on enhancing the model architecture by applying spatial constraints. This contribution is primarily encapsulated in the source file \verb|nn_model/connection_learning.py|. However, this functionality was not incorporated into the implementation discussed throughout this thesis and remains an optional plugin for potential future model extensions. Additionally, Richard Kraus made several minor contributions to other parts of the codebase, aimed at improving memory and computational efficiency, as well as aiding in the installation process on the MetaCentrum\footnote{\url{https://metavo.metacentrum.cz/}} computational cluster. All contributions can be tracked in the GitHub repository.

The dataset utilized throughout this thesis and the evaluation results are stored on the Wintermute cluster of the CSNG MFF CUNI group, with storage details further specified in the attached documentation.

